A pipeline to analyse time-course gene expression data

### 2020-09-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor \texttt{Group} which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).

@@

In both
situations, we first need to set up a object (a \texttt{Moanin} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin} object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the \texttt{Moanin} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatement effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

However, the distribution of the number of genes found differentially
expressed by considering each time-point independently highlights the
challenges of such approach. We can see that some timepoints have many less
genes found significantly differentially expressed (e.g.~timepoint 6H and 18H
of the Kawasaki strain). While there may be biological differences at those
time points for some genes, it seems unlikely that the large majority of genes
differentially expressed at timepoint 3H stop being differentially expressed
at 6H and then jump back to being differentially expressed at 9H. A more
likely explanation is that there is are some technical or biological artifact
about the samples for 6H that is creating higher variation and thus less
ability to detect significance.

@@

In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
the time-course data, for example, overall change versus the largest change.

@@

With a single measure of log-fold change and the p-value, we can now look
at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the ``timecourse'' method of calculating log fold change.

@@

To do this, we first
aggregate all p-values obtained during the time-course differential expresison
step in a single p-value using Fisher's method \citep{fisher:statistical} (\texttt{pvalues\_fisher\_method}).
Next we
select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold
change of at least two for at least one condition and one time-point.

@@

After filtering, we are left with 5521 genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed,
many genes share a similar gene expression pattern, but on different scale.

@@

As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates. Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.

@@

\textbf{Assigning cluster labels:} We could just assign each gene to a cluster
based on which cluster gave the minimum score. By default,
\texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a
sufficiently low enough goodness-of-fit score. The criteria for being
``sufficiently low'' is based on looking at the distribution of the scores of
all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by
\texttt{splines\_kmeans\_score\_and\_label}). A gene is then assigned to a cluster only
if their best score is above the 50\% percentile of that distribution, with the
remaining genes getting \texttt{NA} as their assignment (this choice can be changed
by \texttt{proportion\_genes\_to\_label}). Note that because the distribution of scores
of genes across \emph{all} clusters are used, this is not equivalent to assigning
50\% of genes to a cluster -- it is possible that all genes are assigned to a
cluster.

@@

Now, let us look more in detail some specific clusters. Cluster 6 seems
particularly interesting: it captures genes with strong differences between
the different influenza treatments and the control, while the control remains
relatively flat.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Elbow method}. First introduced in 1953 by Thorndike \citep{thorndike:who},
  the elbow methods looks at the total within cluster sum of squares
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the ``elbow'' is hard to
  see on real data, where the number of clusters is not clearly defined.
\item
  \textbf{Silhouette method}. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
\item
  \textbf{Stability methods} Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every \(k\) to a small randomization of the data. The user is
  then invited to choose the number of cluster based on a number of similarity
  measures.
\end{itemize}

First, let us run the clustering for all values of \(k\) of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignement (or label) for each gene.

@@

The Elbow method to choose the number of clusters relies on visualization aid
to choose the number of clusters. The method relies on plotting the within
cluster sum of squares (WCSS) as a function of the number of clusters. At some
point, the WCSS will start decreasing more slowly, giving an angle or ``elbow''
in the graph. The number of cluster is chosen at this ``Elbow point.''

@@

Methods that rely on the stability of the clustering results to choose \(k\)
thus ensure that the biological interpretation of the clusters hold with
perturbtation to the data. In addition, simulation where the data is generated
with a well defined \(k\) show that the clustering is more stable for the
correct of number of the clusters.

Most methods method to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an \(n \times n\) matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as each elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column correspond to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for \(k=5\):

@@

The model explorer algorithm \citep{ben-hur:stability} proposes to estimate the
number of clusters exploiting the observation that if the number of clusters
is correct, the clustering results are stable to bootstrap resampling, as
described above. The distribution of similaries between bootstrapped results
for each \(k\) can thus be compared for different values of \(k\) and guide the
user in the choice of number of clusters.

@@

The consensus clustering \citep{monti:consensus} relies on a similar idea but
instead of looking at the cumulative density of similarity measures of
bootstrapped clustering, the authors suggests plotting the cumulative density
of elements of the consensus matrix. We provide a function
\texttt{plot\_cdf\_consensus} in \texttt{moanin} to do this

@@

The stability of the clustering based on the consensus matrix can then be
measured via a single number by looking at the area under the curve: the more
stable the clustering, the closer to 0 or 1 will be the entries of consensus
matrix. The consensus clustering strategy thus suggest at looking at either
the AUC as a function of the number of the clusters or the ``improvement'' in
the AUC as a function of the number of cluster.

@@

The consensus clustering method suggest that the most stable is \(k=2\), which
separates over-expressed genes from under-expressed genes. While it is indeed
a very stable clustering, it does not capture the range of gene expression
patterns present in the data. This shows the limitation of such method on real
data, where the number of clusters is not clearly defined.

@@

This workflow provides a tutoral for the analysis of lengthy time-course gene
expression data in R using the package \texttt{moanin}, which aids in implementing
common timecourse analyses. We illustrate the workflow through the analysis of
mice lung tissue exposed to different influenza strain and measured over time.
The proposed workflow covers consists of three common analysis main steps
generally performed after quality control and normalization: (1) differential
expression analysis; (2) clustering of time-course gene expression data; (3)
downstream analysis of clusters. We demonstrate how the use of the package
\texttt{moanin} allows for easy implementation of these procedures in the setting of
time-course data.

@@

All packages use in the workflow are available on GitHub, CRAN, or
Bioconductor.

>>>>

Before we dive into the exploratory analysis and quality control, let us
define color schemes for our data that we will use across the whole analysis.
We define color schemes for groups and time points as named vectors. We also
define a series of markers (or plotting symbols) to distinguish replicate
samples in scatter plots. We also reorder the factor \texttt{Group} which describes
the treatments so that the treatments are ordered from low to high pathogeny
(with Control being first).

@@

In both situations, we first need to set up an object (a \texttt{Moanin} object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin}
object will contain information used throughout this analysis, in particular
the condition and timepoints of each sample and the basis matrix for
fitting splines models.

We start by creating the \texttt{Moanin} object using the \texttt{create\_moanin\_model}
function. We need to provide two things to the function: a \texttt{data.frame} with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata \texttt{data.frame} object should contain at least
two columns: one named \texttt{Group}, containing the treatment effect, and a
second one named \texttt{Timepoint} containing the timepoint information.

@@

However, the distribution of the number of genes found differentially
expressed by considering each time-point independently highlights the
challenges of such approach. We can see that some timepoints have many less
genes found significantly differentially expressed (e.g.~timepoint 6H and 18H
of the Kawasaki strain). While there may be biological differences at those
time points for some genes, it seems unlikely that the large majority of genes
differentially expressed at timepoint 3H stop being differentially expressed
at 6H and then jump back to being differentially expressed at 9H. A more
likely explanation is that there are some technical or biological artifacts
about the samples for 6H that are creating higher variation and thus less
ability to detect significance.

@@

In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary
of each of these methods against each other. We see that each method captures
different elements of the time-course data, for example, overall change versus
the largest change.

@@

With a single measure of log-fold change and the p-value, we can now look at
the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example
of a volcano plot for the comparison of the control to the Kawasaki strain,
using the ``timecourse'' method of calculating log fold change.

@@

To do this, we first
aggregate all p-values obtained during the time-course differential expression
step in a single p-value using Fisher's method \citep{fisher:statistical} (\texttt{pvalues\_fisher\_method}).
Next we
select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold
change of at least two for at least one condition and one time-point.

@@

After filtering, we are left with 5521 genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed,
many genes share a similar gene expression pattern, but on different scales.

@@

As we can see, while these genes have some similarity with the pattern of the
cluster centroids, these particular genes are not the best examples of the
cluster, in the sense of matching the centroid estimates. Because of this
variability in how well the genes fit a cluster, we would like to be able to
score how well a gene fits a cluster.

@@

\textbf{Assigning cluster labels:} We could just assign each gene to a cluster
based on which cluster gave the minimum score. By default,
\texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a
sufficiently low enough goodness-of-fit score. The criteria for being
``sufficiently low'' is based on looking at the distribution of the scores of
all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by
\texttt{splines\_kmeans\_score\_and\_label}). A gene is then assigned to a cluster only
if their best score is above the 50\% percentile of that distribution, with the
remaining genes getting \texttt{NA} as their assignment (this choice can be changed
by \texttt{proportion\_genes\_to\_label}). Note that because the distribution of scores
of genes across \emph{all} clusters is used, this is not equivalent to assigning
50\% of genes to a cluster -- it is possible that all genes are assigned to a
cluster.

@@

Now, let us look more in detail at some specific clusters. Cluster 6 seems
particularly interesting: it captures genes with strong differences between
the different influenza treatments and the control, while the control remains
relatively flat.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Elbow method}. First introduced in 1953 by Thorndike \citep{thorndike:who},
  the elbow method looks at the total within cluster sum of squares
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the ``elbow'' is hard to
  see on real data, where the number of clusters is not clearly defined.
\item
  \textbf{Silhouette method}. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
\item
  \textbf{Stability methods} Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every \(k\) to a small randomization of the data. The user is
  then invited to choose the number of clusters based on a number of similarity
  measures.
\end{itemize}

First, let us run the clustering for all values of \(k\) of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignment (or label) for each gene.

@@

The Elbow method to choose the number of clusters relies on visualization aid
to choose the number of clusters. The method relies on plotting the within
cluster sum of squares (WCSS) as a function of the number of clusters. At some
point, the WCSS will start decreasing more slowly, giving an angle or ``elbow''
in the graph. The number of clusters is chosen at this ``Elbow point.''

@@

Methods that rely on the stability of the clustering results to choose \(k\)
thus ensure that the biological interpretation of the clusters hold with
perturbations to the data. In addition, simulation where the data is generated
with a well defined \(k\) show that the clustering is more stable for the
correct of number of the clusters.

Most methods method to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an \(n \times n\) matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as all elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

@@

Each column corresponds to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for \(k=5\):

@@

The model explorer algorithm \citep{ben-hur:stability} proposes to estimate the
number of clusters exploiting the observation that if the number of clusters
is correct, the clustering results are stable to bootstrap resampling, as
described above. The distribution of similarities between bootstrapped results
for each \(k\) can thus be compared for different values of \(k\) and guide the
user in the choice of number of clusters.

@@

The consensus clustering \citep{monti:consensus} relies on a similar idea but
instead of looking at the cumulative density of similarity measures of
bootstrapped clustering, the authors suggest plotting the cumulative density
of elements of the consensus matrix. We provide a function
\texttt{plot\_cdf\_consensus} in \texttt{moanin} to do this

@@

The stability of the clustering based on the consensus matrix can then be
measured via a single number by looking at the area under the curve: the more
stable the clustering, the closer to 0 or 1 will be the entries of consensus
matrix. The consensus clustering strategy thus suggests at looking at either
the AUC as a function of the number of the clusters or the ``improvement'' in
the AUC as a function of the number of clusters.

@@

The consensus clustering method suggest that the most stable is \(k=2\), which
separates over-expressed genes from under-expressed genes. While it is indeed
a very stable clustering, it does not capture the range of gene expression
patterns present in the data. This shows the limitation of such methods on real
data, where the number of clusters is not clearly defined.

@@

This workflow provides a tutorial for the analysis of lengthy time-course gene
expression data in R using the package \texttt{moanin}, which aids in implementing
common timecourse analyses. We illustrate the workflow through the analysis of
mice lung tissue exposed to different influenza strains and measured over time.
The proposed workflow consists of three common analysis main steps
generally performed after quality control and normalization: (1) differential
expression analysis; (2) clustering of time-course gene expression data; (3)
downstream analysis of clusters. We demonstrate how the use of the package
\texttt{moanin} allows for easy implementation of these procedures in the setting of
time-course data.

@@

All packages used in the workflow are available on GitHub, CRAN, or
Bioconductor.
